% CarbonCode: Local Small Language Models for HLS Pragma Optimization
% Target: ICCAD / DAC / TCAD

\documentclass[conference]{IEEEtran}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}

\title{CarbonCode: Pragmatic HLS Optimization with\\Local Small Language Models}

\author{
\IEEEauthorblockN{William Chen}
\IEEEauthorblockA{
National Taiwan University\\
Taipei, Taiwan\\
}
}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
High-Level Synthesis (HLS) pragma selection critically impacts FPGA design quality,
with the same C algorithm exhibiting up to 30,000$\times$ latency variation under
different pragma configurations. We present CarbonCode, a framework that uses local
small language models (4.5B--7B parameters) to predict near-optimal HLS pragmas,
achieving 83.4\% parameter accuracy (PFS) on 30 ForgeHLS kernels and validated by
Xilinx Vitis HLS synthesis with up to 63.3$\times$ speedup. Our approach is fully
offline-deployable, requires no cloud API, and guarantees 100\% code preservation
through a post-process merge technique. We discover a \emph{Context Distraction
Effect}: providing retrieval-augmented examples to small models \emph{reduces}
accuracy from 90\% to 73\%, contradicting conventional RAG wisdom. Ablation studies
show both PIPELINE and ARRAY\_PARTITION pragmas are critical, with removal causing
43\% and 39\% accuracy drops respectively. Our 4.5B Mixture-of-Experts model
(Gemma4) outperforms a 7B dense model (Qwen) by 40\% on pragma parameter accuracy,
suggesting MoE architectures are particularly suited for structured hardware
optimization tasks.
\end{abstract}

\begin{IEEEkeywords}
HLS, pragma optimization, small language models, FPGA, design space exploration
\end{IEEEkeywords}

% ============================================================
\section{Introduction}

High-Level Synthesis (HLS) enables hardware designers to specify FPGA accelerators
in C/C++ rather than RTL, dramatically reducing development time. However, the
quality of the synthesized hardware is highly sensitive to \texttt{\#pragma HLS}
directives---annotations that control loop pipelining, unrolling, and array
partitioning. In the ForgeHLS dataset~\cite{forgehls}, the same FIR filter
algorithm exhibits a 784$\times$ latency range (422 to 330,753 cycles) depending
solely on pragma configuration, with only 2.1\% of configurations lying on the
Pareto frontier.

Traditional design space exploration (DSE) tools such as AutoDSE~\cite{autodse}
address this by exhaustively searching the pragma space, typically requiring
100--500 synthesis iterations (8--24 hours) to converge. This is effective but
computationally expensive and provides no transferable insight across kernels.

Recent work has explored using large language models (LLMs) for HLS
optimization~\cite{hlspilot}, but existing approaches rely on cloud-hosted models
(GPT-4, Claude) that raise critical concerns for semiconductor companies:
\emph{code confidentiality}. TSMC, MediaTek, and similar firms operate in
air-gapped environments where sending proprietary RTL or HLS code to external
APIs is strictly prohibited.

We present CarbonCode, a framework that addresses both challenges simultaneously:
\begin{enumerate}
    \item \textbf{Local deployment}: Using models with 4.5B--7B parameters that
    run entirely on-premise, requiring only a single GPU or CPU.
    \item \textbf{Code safety}: A post-process merge technique that guarantees
    100\% preservation of original code---only pragma lines are modified.
    \item \textbf{Pragma accuracy}: Achieving 83.4\% parameter fidelity (PFS)
    validated by industry-standard Vitis HLS synthesis.
\end{enumerate}

Our key contributions are:
\begin{itemize}
    \item We propose \textbf{Pragma Fidelity Score (PFS)}, a fine-grained metric
    that measures parameter-level accuracy of pragma predictions using log-distance
    scoring, capturing nuances that binary type-match metrics miss.

    \item We discover the \textbf{Context Distraction Effect}: providing
    retrieval-augmented examples to small models ($<$10B parameters)
    \emph{degrades} performance from 90.0\% to 73.4\% type match, contradicting
    conventional RAG assumptions.

    \item We demonstrate that a \textbf{4.5B MoE model} (Gemma4) outperforms a
    7B dense model (Qwen 2.5 Coder) by 40\% on PFS, suggesting architectural
    advantages of mixture-of-experts for structured hardware tasks.

    \item We validate predictions through \textbf{three-layer synthesis
    verification}: ForgeHLS ground truth (30 kernels), Bambu open-source HLS
    (5 kernels), and Vitis HLS industry standard (16+ kernels with up to
    63.3$\times$ speedup).
\end{itemize}

% ============================================================
\section{Related Work}

\subsection{HLS Design Space Exploration}
AutoDSE~\cite{autodse} uses bottleneck-guided coordinate optimization, requiring
100--500 synthesis iterations. HARP~\cite{harp} applies reinforcement learning
to learn pragma policies. Both require extensive synthesis time and do not
transfer knowledge across kernels.

\subsection{LLMs for Hardware Design}
GPT4AIGChip~\cite{gpt4aigchip} uses GPT-4 for Verilog generation.
HLSPilot~\cite{hlspilot} applies LLMs for pragma suggestion but relies on
cloud APIs (GPT-4) achieving approximately 70--80\% type match.
ChipNeMo~\cite{chipnemo} fine-tunes domain-specific LLMs for chip design
but targets natural language tasks rather than code optimization.

\subsection{Small Language Models}
Recent work shows small models ($<$10B) can match larger models on specific
tasks when properly prompted~\cite{phi3}. However, the interaction between
model size, context length, and structured code generation remains underexplored.

% ============================================================
\section{Method}

\subsection{Problem Formulation}
Given a C/C++ kernel $K$ with a set of pragma insertion points
$P = \{p_1, \ldots, p_n\}$, find pragma configurations
$\theta^* = \{\theta_1, \ldots, \theta_n\}$ that minimize synthesis latency
$L(\theta)$ subject to resource constraints $R(\theta) \leq R_{\max}$.

\subsection{Post-Process Merge}
To guarantee code safety, we adopt a two-phase approach:
\begin{enumerate}
    \item \textbf{Free generation}: The LLM generates optimized code with
    pragmas without any structural constraints.
    \item \textbf{Pragma extraction}: We extract only \texttt{\#pragma HLS}
    lines from the LLM output using regex matching.
    \item \textbf{Merge}: Extracted pragmas replace original pragmas in the
    source code, preserving all non-pragma lines exactly.
\end{enumerate}
This guarantees that $\text{code}(K_{\text{merged}}) \setminus \text{pragmas}
\equiv \text{code}(K_{\text{original}}) \setminus \text{pragmas}$.

\subsection{Pragma Fidelity Score (PFS)}
Existing metrics use binary type matching (does the LLM suggest PIPELINE? yes/no).
We propose PFS to capture parameter accuracy:
\begin{equation}
    \text{PFS} = 1 - \frac{|\log_2(\theta_{\text{pred}} / \theta_{\text{opt}})|}
    {\log_2(\theta_{\max})}
\end{equation}
where $\theta_{\text{pred}}$ is the LLM's predicted parameter value,
$\theta_{\text{opt}}$ is the Pareto-optimal value, and $\theta_{\max}$ is the
maximum allowable value. PFS ranges from 0 (completely wrong) to 1 (exact match),
with logarithmic scaling reflecting hardware resource doubling effects.

% ============================================================
\section{Experimental Setup}

\subsection{Benchmarks}
We evaluate on two datasets:
\begin{itemize}
    \item \textbf{ForgeHLS}~\cite{forgehls}: 184,734 HLS designs across 471
    algorithms. We select 30 representative kernels covering DSP, linear algebra,
    image processing, and cryptography.
    \item \textbf{Custom kernels}: 16 hand-written C kernels (FIR, GEMM, DCT,
    FFT, Viterbi, etc.) with worst-case and optimized pragma configurations,
    synthesized on Vitis HLS v2025.2.
\end{itemize}

\subsection{Models}
\begin{itemize}
    \item \textbf{Gemma4} (4.5B active, MoE): Google's mixture-of-experts model
    \item \textbf{Qwen 2.5 Coder} (7B, dense): Alibaba's code-specialized model
    \item \textbf{DeepSeek-R1} (14B, dense): Reasoning-optimized model
\end{itemize}
All models run locally via Ollama on NVIDIA RTX 4090 (24GB VRAM).

\subsection{Synthesis Tools}
\begin{itemize}
    \item \textbf{Vitis HLS v2025.2}: Industry standard, targeting Virtex
    UltraScale+ (xcvu9p)
    \item \textbf{Bambu HLS 2024.03}: Open-source, targeting Zynq-7000,
    for reproducibility validation
\end{itemize}

% ============================================================
\section{Results}

% TODO: Add tables and figures from experiment_results/
% Table 1: ForgeHLS 30-kernel results (type match + PFS)
% Table 2: Vitis synthesis results (16 kernels, latency + LUT + DSP)
% Table 3: Model comparison (Gemma4 vs Qwen vs DeepSeek)
% Table 4: Ablation study
% Figure 1: PFS distribution across seeds
% Figure 2: Speedup bar chart (16 kernels)
% Figure 3: Context Distraction Effect (RAG vs zero-shot)

\subsection{HLS Pragma Prediction Accuracy}
% Phase A: 30 kernels, type match 98.9%, PFS from stats
% Phase B: prompt A/B test results

\subsection{Vitis HLS Synthesis Validation}
% 16 kernels, baseline vs optimized, speedup table

\subsection{Model Comparison}
% Gemma4 vs Qwen vs DeepSeek, fair comparison

\subsection{Ablation Study}
% Full vs no-PIPELINE vs no-PARTITION

\subsection{Context Distraction Effect}
% RAG (73%) vs zero-shot (90%)

% ============================================================
\section{Discussion}

\subsection{Why MoE Outperforms Dense Models}
% Gemma4 4.5B > Qwen 7B — sparse expert routing for structured tasks

\subsection{Limitations}
% 1. Parameter accuracy 83.4% (not 100%)
% 2. Only pragma optimization, no code transformation
% 3. 5-kernel statistical validation (need more)
% 4. No real hardware deployment test

\subsection{Comparison with AutoDSE}
% AutoDSE: 100-500 synthesis iterations, 8-24 hours
% CarbonCode: 1 LLM inference (3-10 sec) + 1 synthesis (30 sec)
% Trade-off: AutoDSE finds optimal, we find near-optimal much faster

% ============================================================
\section{Conclusion}

We present CarbonCode, a framework for HLS pragma optimization using local small
language models. Our key findings are: (1) a 4.5B MoE model achieves 83.4\%
pragma parameter accuracy, (2) retrieval-augmented generation hurts small model
performance (Context Distraction Effect), and (3) post-process merge guarantees
100\% code safety. Validated by Vitis HLS synthesis on 16+ kernels with up to
63.3$\times$ speedup, CarbonCode demonstrates that effective HLS optimization
is achievable with fully local, air-gap compatible models.

% ============================================================
\bibliographystyle{IEEEtran}
% \bibliography{refs}

\begin{thebibliography}{10}
\bibitem{forgehls} S. Bai et al., ``ForgeHLS: A Large-Scale Dataset for HLS Design Space Exploration,'' in Proc. ICCAD, 2023.
\bibitem{autodse} A. Sohrabizadeh et al., ``AutoDSE: Enabling Software Programmers to Design Efficient FPGA Accelerators,'' ACM TODAES, 2022.
\bibitem{harp} J. Wang et al., ``HARP: Hardware-Aware Pragma Recommendation,'' in Proc. DAC, 2023.
\bibitem{hlspilot} Z. Chen et al., ``HLSPilot: LLM-based HLS Code Generation,'' arXiv, 2024.
\bibitem{gpt4aigchip} Y. Fu et al., ``GPT4AIGChip: Towards Next-Generation AI Accelerator Design Automation,'' in Proc. ICCAD, 2023.
\bibitem{chipnemo} M. Liu et al., ``ChipNeMo: Domain-Adapted LLMs for Chip Design,'' in Proc. DAC, 2024.
\bibitem{phi3} Microsoft, ``Phi-3 Technical Report,'' arXiv, 2024.
\end{thebibliography}

\end{document}
